---
title: "backupGov50projRMD"
author: "Arjun Akwei"
date: "12/11/2022"
output: html_document
---

---
title: "Gov 50 Final Project"
author: "Arjun Akwei"
description: "The Effect of EU Membership on Economic Development"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

For this project I explored how determining whether membership in a liberal international organization such as the EU has caused new member countries to develop economically in a significant way. Between 1990 and 2005 several former Soviet sphere countries were making the decision of whether or not to join the EU, and many EU members and allies listed greater economic development as a key selling point of joining. I was curious to assess whether such an impact was really seen. 

My project curated a sample of countries from the "Eastern Bloc" that joined the EU in the same year, and mapped years spent in EU against their annual scores on the UN's Human Development Index. I then conduct a difference in difference analysis using a sample of countries from that bloc that chose not to join the EU as a control group to determine the economic treatment effect of EU membership.

**Sample of Eastern Bloc countries now in EU**  
- Poland 2004  
- Czech Republic 2004  
- Estonia 2004  
- Hungary 2004  
- Slovakia 2004  
**Sample of Eastern bloc countries NOT in EU**  
- Albania  
- Moldova  
- Ukraine  
- Belarus  
- Georgia  

## Hypothesis and discussion of data
I hypothesized, facilitated integration into the global economy, as well as direct support from the EU itself, would cause member organizations to develop at rates far faster than they would have without joining. The explanatory variable in this case is time spent as a member of the EU, something measured in years since being granted membership in 2004. A separate set of countries from the same region that did NOT join the EU will be established as a comparative to isolate the difference. The outcome variable I will be tracking is economic development, as quantified by the UN's Human Development Index. This is an annual evaluation of how developed a given country is. It incorporates the following factors: life expectancy at birth, mean of years of schooling for adults aged 25 years and more, expected years of schooling for children of school entering age, and gross national income per capita. The data is published for each country as far back as 1990, but I had to combine these country data-sets into one in order to perform my analysis.

Rather than simply assessing growth since joining, I chose to conduct a difference in differences analysis to isolate the impact of EU membership from the impact of year on year growth. I expected that over time most of these countries will have trended towards becoming more developed, but a difference-in-differences analysis will reveal whether the explanatory had some additional effect. If the difference-in-differences analysis indicates a higher rate of development in member countries vs non-member countries, this will suggest that time spent within the EU is positively related to development, proving the hypothesis correct. No significant difference or a negative relationship would disprove the hypothesis. 


```{r include= FALSE}
library(tidyverse)
Albania <- read_csv("Albania.csv")
Belarus <- read_csv("Belarus.csv")
Czechia <- read_csv("Czechia.csv")
Estonia <- read_csv("Estonia.csv")
Georgia <- read_csv("Georgia.csv")
Hungary <- read_csv("Hungary.csv")
Moldova <- read_csv("Moldova (Republic of).csv")
Poland <- read_csv("Poland.csv")
Slovakia <- read_csv("Slovakia.csv")
Ukraine <- read_csv("Ukraine.csv")

```

```{r slicing, include= FALSE}

Albania_slice <- Albania |>
  slice(6:37) |>
  mutate(HDI_index = value,
         country = "Albania") |>
  select(c(key, HDI_index, country))
Albania_slice

country_slice <- function(country) {
  country_slice <- slice(country, 6:37) |>
    mutate(HDI_index = value) |>
    select(c(key, HDI_index))
  return(country_slice)
}

Belarus_slice <- country_slice(Belarus) |>
  mutate(country = "Belarus")
Czechia_slice <- country_slice(Czechia) |>
  mutate(country = "Czechia")
Estonia_slice <- country_slice(Estonia) |>
  mutate(country = "Estonia")
Georgia_slice <- country_slice(Georgia) |>
  mutate(country = "Georgia")
Hungary_slice <- country_slice(Hungary) |>
  mutate(country = "Hungary")
Moldova_slice <- country_slice(Moldova) |>
  mutate(country = "Moldova")
Poland_slice <- country_slice(Poland) |>
  mutate(country = "Poland")
Slovakia_slice <- country_slice(Slovakia) |>
  mutate(country = "Slovakia")
Ukraine_slice <- country_slice(Ukraine) |>
  mutate(country = "Ukraine")
```

```{r joining, include= FALSE}
two_count <- rbind(Albania_slice, Belarus_slice)
two_count

HDI <- rbind(two_count, Czechia_slice, Estonia_slice, Georgia_slice, Hungary_slice, Moldova_slice, Poland_slice, Slovakia_slice, Ukraine_slice)
HDI

```

```{r classifying, include= FALSE}
HDI$HDI_index <- as.numeric(as.character(HDI$HDI_index))

HDI_classified <- HDI |>
  mutate(EU_member = if_else(country %in% c("Poland", "Czechia", "Estonia", "Hungary", "Slovakia"), "yes","no"),
         year = substring(key, 34,37)) |>
  mutate(EU = if_else(country %in% c("Poland", "Czechia", "Estonia", "Hungary", "Slovakia"), 1,0),
         year = substring(key, 34,37)) |>
  group_by(country,EU) |>
  select(year, country, HDI_index, EU, EU_member)
HDI_classified

```
## Graph 1: Country HDI 1990-2021 (categorized by EU membership)
```{r fig.width = 11, fig.height = 4, echo= FALSE}
comparative_graph <- HDI_classified |>
  ggplot(aes(x = year, y = HDI_index, color = EU_member)) +
  geom_line(aes(group=country)) +
  labs(x = "Year", y = "HDI", title = "Comparative HDI Growth")

comparative_graph
  

```

## Graph 2: Aggregated HDI scores over time (EU vs Non-EU)
```{r, include=FALSE}
comparative_analysis <- HDI_classified |>
  group_by(year, EU_member, EU) |>
  summarize(mean_HDI = mean(HDI_index, na.rm = TRUE))
comparative_analysis
```


```{r fig.width = 11, fig.height = 3, echo=FALSE}
avg_plot <- comparative_analysis |>
  ggplot(aes(x = year, y = mean_HDI, color = EU_member)) +
  geom_line(aes(group=EU_member)) +
  labs(x = "Year", y = "HDI", title = "Comparative HDI Growth")
avg_plot
```

## Regression: Difference in Differences
```{r did, echo = FALSE}
#Diff in Diff
comparative_analysis$time = ifelse(comparative_analysis$year >=2004, 1, 0)
comparative_analysis$did = comparative_analysis$time * comparative_analysis$EU

didreg = lm(mean_HDI ~ EU + time +did, data = comparative_analysis)
summary(didreg)
```


## Results and Analysis
To test my hypothesis I used a difference in differences analysis, assuming that since all of the countries were from the Soviet aligned Eastern Bloc they made good peers. This is supported by their relatively convergent HDI scores pre-2004. To conduct this analysis, several steps were taken. First, I created a dummy variable that would reflect treatment starting at 2004. Second I created a did variable that is the interaction between time and treatment which would be "did" or difference in differences.From there I ran the regression, and the coefficient for did  is the differences-in-differences estimator. This would isolate the impact that EU membership made for growth each year, removing the impacts of the other two variables in the regression ("time", representing general year on year progress, and "EU" representing the different starting positions in 2004).

Based on the difference in differences analysis I ran, the coefficients yielded for did were 0.0007643 and 0.0123717 for the estimate and standard error respectively. This estimator suggests that EU members year-on-year saw an additional 0.076% growth in HDI compared to the control group of non-EU countries. This would suggest that EU membership does have a positive, though slight, impact on HDI growth beyond what would be seen in normal year-on-year progress. The standard error for the did estimator is 1.2%. Though this error is large relative to the statistic measured itself, the p-value given in the summary of the regression is <2.2 e-16, which is well below the generic alpha level of 0.05. This would mean that we are able to reject a null hypothesis that there was no significant difference in HDI growth between the treatment and control groups, meaning the results are in fact statistically significant. However, this impact is significantly smaller than the time coefficient, suggesting that year-over-year progress regardless of EU membership was the bigger driver of HDI growth.

The difference in differences analysis strengthens the argument for that effect being causal rather than just correlational because the analysis isolated EU impact from year-on-year change that is the primary confounder. This shows that EU membership does in fact lead to greater economic growth.

## Conclusion
To circle back to the initial question posed (whether EU membership accelerated growth rates for its members compared to non-members), it seems the promises made by the EU rang true, though perhaps were over-exaggerated. Ultimately the EU did increase rates of HDI growth, just by a marginal amount. This operationalization of growth may be too narrow to fully demonstrate the returns of joining the EU though-- there is plenty of evidence to suggest that EU membership improved respect for civil liberties relative to non-members, as well as provided support in times of economic crisis (displayed partially by the smaller dip in HDI during the 2008-2009 global recession). Thus, there may yet be an even greater impact than was displayed here.

This study was limited in that the countries I chose as the control group were imperfect foils to the treatment group. Though their HDIs were closer pre 1991, there was already a difference in HDI to start that may have impacted my findings. Perhaps if compared to a more similar control, different results would have been found. Thus, one meaningful improvement to make would be doing a more sophisticated selection of countries beyond my sorting of Soviet Eastern Bloc countries.






